A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-26 natural history of scavenging.tex
moved around sections
paper is effectively encounter rate and handling time

\subsection*{The Challenges of Scavenging} 
The chief hurdle to scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some opportunistic scavenging. 
However, species that rely on scavenging to sustain substantial portions of their diets must encounter a sufficient amount of carrion in order to meet their energetic demands.
Once found, the scavenger must be able to out-compete any potential competitors and process the, typically decaying, carcass replete with microorganism derived toxins  \citep{ruxton2014fruit}. 
Both of these characteristics can be assumed to be under evolutionary selection pressures for traits that increase carrion discovery and monopoly.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics and are outside the selection pressures on the scavenger.
Each of these facets are essentially the backbone of fundamental ecological theory and are the key parameters defined in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}.  
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place any vertebrate species, past or present, and assess the importance of carrion in its diet.

@@

%KH: need paragraph for how do we know something is (was) a scavenger
\section*{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
Locomotory speed, foraging time and detection radius all increase the encounter rate between a scavenger and the carcasses it is searching for. 
We would thus expect selection pressures to act on the various traits that govern these parameters.  
% However, encounter rate is also determined by the productivity of the environment, something which the selective forces acting on the scavenging have no control over.

@@

\subsection*{Locomotion}
Because of the inherent unpredictability of carrion, scavenging depends more on the ability to efficiently move over large areas than predation.
This generally requires an efficient transfer of metabolic energy into movement which relies on the animal's anatomy and physiology as well as the medium of the environment in which the animal is moving (i.e. aerial, aquatic or terrestrial).
Perhaps the most efficient form of locomotion in vertebrates is, paradoxically, found in flying species. 
Despite the energetic costs of flight, the only known vertebrate obligate scavengers are the old and the new world vultures. 
While powered flight is energetically expensive, species like vultures have evolved behavioural and anatomical features to exploit air currents using their large wingspans, allowing them to soar at a cost of only twice their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}.
By depending on thermal air flows these species can forage over vast ranges \citep{spiegel2013factors}. 
An analogous mode of locomotion is also exploited by seabirds, who use strong ocean winds to search large areas of the oceans \citep{norberg2012vertebrate,thaxter2012seabird}. 
While many species of seabird are likely primarily predators, it seems that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. 
This is typically in the form of squid carcases, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

The groups from which these modern soaring birds arose, appear during the Palaeocene \citep[66 - 56 Million years ago (Mya); ][]{Jetz2012, Jarvis2014} and Cretaceous \citep{chiappe2006early} respectively. 
However, soaring flight is likely to be far older than this with avian flight originating in the Late Jurassic (163.5-145 Mya) and vertebrate flight in the Late Triassic (235-201.3 Mya) coincident with the pterosaurs. 
Indeed, scavenging among pterosaurs has been hypothesised many times before \citep{witton2008reappraisal}. 
Certain groups of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
It seems probable that as least some of these extinct species used soaring as a means for scavenging \citep{witton2013pterosaurs}. 
%KH added "at least some of these" as it was not clear what species used soaring as a means for scavenging.
While soaring is perhaps the only viable means of locomotion that allows for an obligate, scavenging life-style \citep{ruxton2004obligate}, powered flight is still an efficient means of locomotion. 
Certainly, avian flight is cheaper than either walking or running \citep{tucker1975energetic}.

@@

Sharks are one likely candidate for general scavenging behaviors as their locomotion, which depends on large pectoral fins to generate lift as they swim, resembles that of the large soaring fliers.
%changed this sentence up a bit 
Many shark species have large foraging ranges \citep[e.g. the great white sharks \textit{Carcharodon carcharias};][]{bruce2006movements} and it seems reasonable that they would use oceanographic currents to further reduce movement costs \citep{ruxton2004energetic}. %and are able to maintain a type of endothermy from thier movements \citep{bernal2015sharks}. %TG: I guess the following is a comment: (I know the Greenland shark is the slowest swimmer). 
In fact, facultative scavenging is seen in many selachian groups, including species of extant sharks like white sharks \citep[known to feed on whale carcasses;][]{fallows2013white}, Greenland sharks \citep[feeding on seals;][]{watanabe2012slowest}, and sixgill sharks \citep{anderson2016impact}. 
There is evidence too of scavenging in extinct species, where shark teeth have been found in the remains of dinosaurs, mosasaurs and Pliocene mysticete whales \citep[5.3-3.6 Mya; ][]{schwimmer1997scavenging,ehret2009caught}.

Interestingly, style of swimming in fish does not significantly affect the cost of movement \citep{williams1999evolution}, hence it is likely that many aquatic species with large ranges will encounter scavenging opportunities. 
It is perhaps jaw morphology that is more likely to define which species can avail of carcasses.
%KH I moved this down here as it kinda throws our shark soaring thing out the window (which I like as a linking function). However I think this kinda solves our problems in the complexity of swimming by saying apart from large ranges ability to eat is probably more important.
%KH also is fish an ok term to use, will people get pissy or anything like with reptiles?

% KH endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism marine mammals would especially prosper as scavengers. %, especially as this group has perhaps the largest home range of any animals (whales and shit).
Fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their evolutionary trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, the total energetic cost is similar to an equivalent terrestrial or aerial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an energetically expensive system. 
That said, aquatic endotherms have and do scavenge. 
For instance, early whales such as \textit{Basilosaurus} (38-36.5 Mya) seem to have fit into the same niche as killer whales (\textit{Orcinus orca}) and we have some evidence for scavenging in both \citep{fahlke2012bite,Whitehead415}.

@@

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move \citep{tucker1975energetic}. Unlike aerial and aquatic environments support must be provided through the animals posture leading to trade offs between low cost of maintain and the efficiency of transport. For example, the sprawling posture of many ectotherm groups (except frogs!) would be expected to allow for a lower maintenance cost relating to body support in comparison to groups with the limbs position underneath the body, such as mammals, aves and dinosaurs.
%KH: I agree with TG and have removed the gas exchange thing as its a physiology thing only applicable to certain groups and not got to do with terrestrial environments really. I also removed the 
%Also that reference is from 1975, is this up to date?

%
%TG: It's missing a transition here, we go from gas exchanges to posture. Maybe we need something like "Among terrestrial scavengers, most are mammals. This might be due to the posture blablabalba." But then it doesn't connect logically with the sentence before: "mammals have are inefficient...". Maybe one way is to just remove the gas exchange and start with something like:
%Terrestrial environments are the most energetically costly in which to move, which may be due to the low muscular efficiency of running \citep{tucker1975energetic}.
%Most terrestrial species that have been observed (or are suspected to be) scavenging are mammals.
%TG: and then link to the posture hypothesis (with the "meh, we're not sure" at the end).
%
While the low maintenance costs of a sprawling like posture may reduce the need to scavenge frequently, the ability to search large areas is several restricted in these groups hence likely reducing such opportunities (ref on home range of terrestrial ectoterms). For example, a sprawling gait can results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. On the other hand the more erect body positions and often associated higher basal metabolic rates, allow for sustained locomotion over long periods of time and hence allow such species to search large areas.

The early transition from a sprawling gait, such as seen in \textit{Tiktaalik}, to the more erect posture of Synapsids and later dinosaurs and mammals, has often been supposed as conferring a huge advantage to the latter.
%KH Even after changing it I dont like this transition type narrative.
%KH also needs to be more clear on what the advantage to scavenging is for these groups.
%The purported advantages include benefits in terms of speed, efficiency, muscle effort and manoeuvrability \citep{sullivan2015posture}. 
%none of these advantages refer very clearly to scavenging
Despite being intuitive, \cite{sullivan2015posture} states most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
%KH I'm always wary of deeming something as intuitive,

@@

%This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
%This would also have been true of extinct species with the same physiology. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
%Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move. 
Whatever the case, it is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion. 
Although the earliest evidence of vertebrate scavenging comes from the Permian (298.9 - 252.17 Mya) where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}.

Modern endothermic mammals can sustain longer periods of energetically expensive activity \citep{bennett1979endothermy} resulting in larger foraging ranges. 
Today, terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizeable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
And although no contemporary terrestrial vertebrate exists as an obligate scavenger, most, if not all, are facultative to some extent \citep{beasley2015vertebrates}.
The particular reliance of hyenas on carrion means we can use them as examples of efficient terrestrial scavengers to compare with other forms. 
In terms of locomotion, they employ a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently, loping at 10 km/hr \citep{mills1989comparative,jones2015african}. 
Such long-distance travel is apparent in African wild dogs (\textit{Lycaon pictus}) and many other canids \citep{pennycuick1995radius,janis2014forelimb}. 
In contrast, big cats like leopards (\textit{Panthera pardus}) rely on ambush \citep{pennycuick1995radius}. 
This allows us to make a broad distinction between the ambush strategies of cat-like forms and the pursuit/ pounce strategies of more dog-like forms, the latter being more suited to scavenging \citep{janis2014forelimb}. 
These insights allow us to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene (56-33.9 Mya) where the order split into the Caniforma and Feliforma \citep{van1987skeletal}.
To take one example, \cite{anyonge1996locomotor} found that \textit{Nimravides}, a genus of sabretooth cat from the Miocence (10.3 to 5.3 Mya), were likely to have been ambush predators which would argue against them taking a lot of carrion.

Of course, terrestrial animals can also move bipedally. 
Although the evolution of bipedal  movement was significant in that it freed up the forelimbs for other purposes (e.g. climbing, tool-use, wing development etc.) it does not differ radically in cost from quadrupedal locomotion (\cite{williams1999evolution}, and references therein). 
For instance, \cite{alexander2004bipedal} shows  that, in the case of humans, we are more economical than predicted while walking and less so while running according to predicted costs of terrestrial movement.
Our locomotory efficiency has fed into the question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene, which has been a matter of debate for years \citep{dominguez2002hunting}.
\cite{ruxton2013endurance} added to this debate with his argument that long distance endurance running, often used by humans for hunting, was not an important feature of hominin scavenging.
He arrives at this conclusion on the basis of the high cost of running coupled with the fact that dangerous competitors will tend to arrive beforehand.

@@

Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been extensively explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
While the locomotory ability of theropods has been debated since their first inception, more recent studies have reconstructed them as relatively mobile animals \citep{pontzer2009biomechanics}. 
Despite some suggestions that larger species may have had some advantage in scavenging, partially due to the ability to search large areas \citep{ruxton2003could}, more recent studies have shown that the energetic demands of such large forms meant scavenging was likely more prevalent in mid-sized theropods of approximately half a tonne \citep{carbone2011intra,kane2016body}

@@

As predicted by the importance of an increased encounter rate, scavengers have well-developed senses, with the visual and olfactory sensory systems most often associated with scavenging behavior. 
This is perhaps no surprise because sensory systems that rely on detecting signals associated with living animals, such as audioception, electroreception, thermoreception and echolocation will be limited in their ability to detect an already dead animal.

@@

Indeed, the olfactory senses of many extant (and in all probability extinct) carnivores meet this required distance, making scavenging feasible for most terrestrial carnivores \citep{farlow1994speculations,mech2010wolves}. 
Among extinct species in particular, we can use the ratio of olfactory bulb to brain size to infer a preference for olfactory foraging \citep{zelenitsky2011evolution}.
This approach was used by \cite{zelenitsky2011evolution} to hypothesise such a mode for the theropod dinosaur \textit{Bambiraptor} and by \cite{witmer2009new} for tyrannosaurs.
The flying pterosaurs however, had tiny olfactory bulbs indicating this sense was not relied on \citep{witton2013pterosaurs}

Species capable of flight have effectively added an extra spatial dimension (i.e. the vertical component) to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Certainly, vultures are known to have impressive visual acuity, with one estimate indicating lappet-faced vultures (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed vision \citep{reymond1985spatial}.
It follows that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts through vision \citep{AR:AR22815} as the pterosaurs convergently evolved large orbits and optic lobes \citep{witton2013pterosaurs}.

@@

Although aquatic species also have a vertical component because they forage in a water body, they must contend with the low-light environment where visual detection distances are far lower ($<$ 100 m) than they would be for air.
As such, aquatic animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}.
This is particularly relevant to sharks and aquatic snakes who are deemed as having the most suitable physiology for scavenging.
A hypothesis put forth by \cite{sazima1990necrofagia} argued that chemical gradients in water would allow for a relatively easier detection of carrion by snakes.
This gained some support from \cite{devault2002scavenging}, who found a preponderance of aquatic snake species in their review of this behaviour.
Smell seems to be the primary means of carcass detection in sharks as well. 
\cite{fallows2013white} found that wind speed determined the number of sharks feeding at whale carcasses, indicating they were dependent on detecting the odours from the decaying whales.

>>>>

\subsection*{The Challenges of Scavenging} 
The chief hurdle to scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some opportunistic scavenging. 
However, species that rely on scavenging to sustain substantial portions of their diets must encounter a sufficient amount of carrion in order to meet their energetic demands.
Once found, the scavenger must be able to out-compete any potential competitors and process the, typically decaying, carcass replete with micro-organism derived toxins  \citep{ruxton2014fruit}. 
These characteristics can be assumed to be under evolutionary selection pressures for traits that increase carrion discovery and monopoly.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics but are outside the selection pressures on the scavenger.
Each of these facets are essentially the backbone of fundamental ecological theory and are the key parameters defined in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}.  
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place any vertebrate species, past or present, and assess the likely importance of carrion in its diet.

@@

%KH: need paragraph for how do we know something is (was) a scavenger
\section*{Encounter Rate}
All foraging processes depend on the encounter rate between consumer and resource. 
Locomotory speed, foraging time and detection radius all increase the encounter rate between a scavenger and the carcasses it is searching for. 
Alternatively, a species can reduce its metabolic requirements so that it can survive long periods between meals. 
We would thus expect selection pressures to act on the various traits that govern these parameters.

However, as we noted above, encounter rate is also determined by the productivity of the environment, something which the selective forces acting on the scavenging have no control over.

@@

\subsection*{Metabolism}
Because of the sporadic nature of carrion we would expect adaptations that reduce energetic costs of maintenance to be selected for in scavengers as it would allow for longer inter-feeding periods. 
Extant reptiles possess an advantage here, in that over the course of a year their food requirements can be 30 times lower than an endotherm of equal size \citep{Nagy1621}.
\cite{devault2002scavenging} suggest this is an avenue for scavenging in snakes because they ``exhibit exceedingly low maintenance metabolisms, and most can survive on a few scant feedings per year.
It is, therefore, possible for snakes to rely largely on infrequent, less energy-rich meals."
In the same review the authors found occurrences of scavenging spread across five families of snakes and stated that this behaviour is ``far more common than currently acknowledged."\citep{devault2002scavenging}.
The same reasoning can be applied to crocodiles and their allies \citep{forrest2003evidence} because a sit and wait strategy is viable for an ectotherm \citep{forrest2003evidence,moleon2015carcass}. 
\cite{carey1982temperature} found ectothermic sharks have the ability to exist weeks between meals because they focus on the energy-rich sections of carcasses (see Handling Time). 
Endotherms have also evolved physiological mechanisms that allow them to depress their otherwise high metabolic rates at certain times e.g. vultures who do so while resting at their roost and during periods of food deprivation \citep{bahat1998nocturnal}.

\subsection*{Locomotion}
Because of the inherent unpredictability of carrion, scavenging depends more on the ability to efficiently move over large areas than does predation.
This generally requires an efficient transfer of metabolic energy into movement which relies on the animal's anatomy and physiology as well as the medium of the environment in which the animal is moving (i.e. aerial, aquatic or terrestrial).
Perhaps the most efficient form of locomotion in vertebrates is, paradoxically, found in flying species. 
Despite the energetic costs of flight, the only known vertebrate obligate scavengers are the old and the new world vultures. 
While powered flight is energetically expensive, species like vultures have evolved behavioural and anatomical features to exploit air currents using their large wingspans, allowing them to soar at a cost of only twice their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}.
By depending on thermal air flows these species can forage over vast ranges \citep{spiegel2013factors}. 
An analogous mode of locomotion is also exploited by seabirds, who use strong ocean winds to search large areas of the oceans \citep{norberg2012vertebrate,thaxter2012seabird}. 
While many species of seabird are likely primarily predators, it seems that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. 
This is typically in the form of squid carcases, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

The groups from which these modern soaring birds arose, appear during the Palaeocene \citep[66 - 56 Million years ago (Mya); ][]{Jetz2012, Jarvis2014} and Cretaceous \citep{chiappe2006early} respectively. 
However, soaring flight is likely to be far older than this with avian flight originating in the Late Jurassic (163.5-145 Mya) and vertebrate flight in the Late Triassic (235-201.3 Mya) coincident with the pterosaurs. 
Indeed, scavenging among pterosaurs has been hypothesised many times before \citep{witton2008reappraisal}. 
Certain groups of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
It seems probable that at least some of these extinct species used soaring as a means for scavenging \citep{witton2013pterosaurs}. 
%KH added "at least some of these" as it was not clear what species used soaring as a means for scavenging.
While soaring is perhaps the only viable means of locomotion that allows for an obligate, scavenging life-style \citep{ruxton2004obligate}, powered flight is still an efficient means of locomotion. 
Certainly, avian flight is cheaper than either walking or running \citep{tucker1975energetic}.

@@

Sharks are one likely candidate for general scavenging behaviors as their locomotion, which depends on large pectoral fins to generate lift as they swim, resembles that of the large soaring fliers.
Many shark species have large foraging ranges \citep[e.g. the great white sharks \textit{Carcharodon carcharias};][]{bruce2006movements} and it seems reasonable that they would use oceanographic currents to further reduce movement costs \citep{ruxton2004energetic}.
In fact, facultative scavenging is seen in many selachian groups, including species of extant sharks like white sharks \citep[known to feed on whale carcasses;][]{fallows2013white}, Greenland sharks \citep[feeding on seals;][]{watanabe2012slowest}, and sixgill sharks \citep{anderson2016impact}. 
There is evidence too of scavenging in extinct species, where shark teeth have been found in the remains of dinosaurs, mosasaurs and Pliocene mysticete whales \citep[5.3-3.6 Mya; ][]{schwimmer1997scavenging,ehret2009caught}.

Interestingly, style of swimming in fish does not significantly affect the cost of movement \citep{williams1999evolution}, hence it is likely that many aquatic species with large ranges will encounter scavenging opportunities. 
It is perhaps jaw morphology that is more likely to define which species can avail of carcasses.
%KH I moved this down here as it kinda throws our shark soaring thing out the window (which I like as a linking function). However I think this kinda solves our problems in the complexity of swimming by saying apart from large ranges ability to eat is probably more important.

We might expect then that by combining an aquatic environment and an endothermic metabolism marine mammals would especially prosper as scavengers. %, especially as this group has perhaps the largest home range of any animals (whales and shit).
Fossil pinnipeds and cetaceans from 60 Mya have transitional features indicative of their evolutionary trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, the total energetic cost is similar to an equivalent terrestrial or aerial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an energetically expensive system. 
That said, aquatic endotherms have and do scavenge. 
For instance, early whales such as \textit{Basilosaurus} (38-36.5 Mya) seem to have fit into the same niche as killer whales (\textit{Orcinus orca}) and we have some evidence for scavenging in both \citep{fahlke2012bite,Whitehead415}.

@@

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move \citep{tucker1975energetic}. 
Unlike aerial and aquatic environments support must be provided through the animal's posture. % leading to trade offs between low cost of maintain and the efficiency of transport. 
The early transition from a sprawling gait, seen in early tetrapods, to the more erect posture of synapsids and later dinosaurs and mammals, has often been supposed as conferring a huge advantage to the latter groups \citep{sullivan2015posture}.
The purported advantages include benefits in terms of speed, efficiency, muscle effort and manoeuvrability \citep{sullivan2015posture}.
Clearly, for a scavenger, an ability to efficiently cover an area at a high speed would increase the encounter rate with carrion. 
Despite being intuitive, \cite{sullivan2015posture} states most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
One noted consequence of a sprawling gait is the phenomenon known as Carrier's constraint such that the animal can't move and undergo costal ventilation at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
The evolution of an upright posture has been offered as one of the primary mechanisms that allowed early archosaurs to overcome this constraint \citep{uriona2008recruitment}.

% For example, the sprawling posture of many ectotherm groups (except frogs!) would be expected to allow for a lower maintenance cost relating to body support in comparison to groups with the limbs position underneath the body, such as mammals, aves and dinosaurs.
%KH: I agree with TG and have removed the gas exchange thing as its a physiology thing only applicable to certain groups and not got to do with terrestrial environments really. I also removed the 
%Also that reference is from 1975, is this up to date?

% While the low maintenance costs of a sprawling-like posture may reduce the need to scavenge frequently, the ability to search large areas quickly is severely restricted in these groups reducing scavenging opportunities (ref on home range of terrestrial ectoterms). 
% For example, a sprawling gait results in 
% On the other hand the more erect body positions and often associated higher basal metabolic rates, allow for sustained locomotion over long periods of time allowing such species to search large areas.

%KH Even after changing it I dont like this transition type narrative.
%KH also needs to be more clear on what the advantage to scavenging is for these groups.
%
%none of these advantages refer very clearly to scavenging
%KH I'm always wary of deeming something as intuitive,

@@

%This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
%This would also have been true of extinct species with the same physiology. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
%Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move.

% Although the earliest evidence of vertebrate scavenging comes from the Permian (298.9 - 252.17 Mya) where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
Whatever the case, it is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.
Modern endothermic mammals can sustain longer periods of energetically expensive activity \citep{bennett1979endothermy} resulting in larger foraging ranges. 
Today, terrestrial scavenging in the mammals is probably best known in an African context where hyenas, jackals and lions all take sizeable proportions of carrion in their diet.
In the spotted hyena (\textit{Crocuta crocuta}), striped hyena (\textit{Hyaena hyaena}) and brown hyena (\textit{Hyaena brunnea}) it can be over 90\% \citep{jones2015african}.
And although no contemporary terrestrial vertebrate exists as an obligate scavenger, most, if not all, are facultative to some extent \citep{beasley2015vertebrates}.
The particular reliance of hyenas on carrion means we can use them as examples of efficient terrestrial scavengers to compare with other forms. 
In terms of locomotion, they employ a characteristic ``rocking horse gait"  which allows them to cover great distances efficiently, loping at 10 km/hr \citep{mills1989comparative,jones2015african}. 
Such long-distance travel is apparent in African wild dogs (\textit{Lycaon pictus}) and many other canids \citep{pennycuick1995radius,janis2014forelimb}. 
In contrast, big cats like leopards (\textit{Panthera pardus}) rely on ambush \citep{pennycuick1995radius}. 
This allows us to make a broad distinction between the ambush strategies of cat-like forms and the pursuit/ pounce strategies of more dog-like forms, the latter being more suited to scavenging \citep{janis2014forelimb}. 
These insights allow us to compare extant terrestrial species to their prehistoric forebears given the dominance of mammalian carnviores since the Eocene (56-33.9 Mya) where the order split into the Caniforma and Feliforma \citep{van1987skeletal}.
To take one example, \cite{anyonge1996locomotor} found that \textit{Nimravides}, a genus of sabretooth cat from the Miocence (10.3 to 5.3 Mya), were likely to have been ambush predators which would argue against them taking a lot of carrion.

Of course, terrestrial animals can also move bipedally. 
Although the evolution of bipedal  movement was significant in that it freed up the forelimbs for other purposes (e.g. climbing, tool-use, wing development etc.) it does not differ radically in cost from quadrupedal locomotion (\cite{williams1999evolution}, and references therein). 
For instance, \cite{alexander2004bipedal} shows  that, in the case of humans, we are more economical than predicted while walking and less so while running according to predicted costs of terrestrial movement.
Our locomotory efficiency has fed into the question of where our ancestors placed on the hunter-scavenger axis during the Plio-Pleistocene, which has been a matter of debate for years \citep{dominguez2002hunting}.
\cite{ruxton2013endurance} added to this debate with their argument that long distance endurance running, often used by humans for hunting, was not an important feature of hominin scavenging.
They arrive at this conclusion on the basis of the high cost of running coupled with the fact that dangerous competitors will tend to arrive beforehand.

@@

Aside from humans and our allies, the best-known terrestrial bipeds are the dinosaurs and unsurprisingly, given their enduring appeal, the prevalence of scavenging has been extensively explored in the carnivorous theropods.
These were the dominant terrestrial carnivores for most of the Mesozoic Era (252.17 - 66 Mya) and ranged from the chicken-sized to the whale-sized, all of which were bipedal.
While the locomotory ability of theropods has been debated since their first inception, more recent studies have reconstructed them as relatively mobile animals \citep{pontzer2009biomechanics}. 
Despite some suggestions that larger species may have had some advantage in scavenging, partially due to the ability to search large areas \citep{ruxton2003could}, more recent studies have shown that the energetic demands of such large forms meant scavenging was likely more prevalent in mid-sized theropods of approximately half a tonne \citep{carbone2011intra,kane2016body}.

@@

As predicted by the importance of an increased encounter rate, known scavengers have evolved well-developed senses, with the visual and olfactory sensory systems most often associated with scavenging behavior. 
This is perhaps no surprise because sensory systems that rely on detecting signals associated with living animals, such as audioception, electroreception, thermoreception and echolocation will be limited in their ability to detect an already dead animal.

@@

Indeed, the olfactory senses of many extant (and in all probability extinct) carnivores meet this required distance, making scavenging feasible for most terrestrial carnivores \citep{farlow1994speculations,mech2010wolves}. 
Among extinct species in particular, we can use the ratio of olfactory bulb to brain size to infer a preference for olfactory foraging \citep{zelenitsky2011evolution}.
This approach was used by \cite{zelenitsky2011evolution} to hypothesise such a mode for the theropod dinosaur \textit{Bambiraptor} and by \cite{witmer2009new} for tyrannosaurs.
The flying pterosaurs however, had tiny olfactory bulbs indicating this sense was not relied on \citep{witton2013pterosaurs}.

Species capable of flight have effectively added an extra spatial dimension (i.e. the vertical component) to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Certainly, vultures are known to have impressive visual acuity, with one estimate indicating lappet-faced vultures (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed vision \citep{reymond1985spatial}.
The flying pterosaurs also convergently evolved large orbits and optic lobes \citep{witton2013pterosaurs}. 
It follows that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts through vision \citep{AR:AR22815}.

@@

Although aquatic species also have a vertical component to their environment, they must contend with low-light levels where visual detection distances are far lower ($<$ 100 m) than they would be for air.
As such, aquatic animals detect resources through chemo- and mechanoreception more so than through vision \citep{ruxton2004energetic}.
This is particularly relevant to sharks and aquatic snakes who are deemed as having the most suitable physiology for scavenging.
A hypothesis put forth by \cite{sazima1990necrofagia} argued that chemical gradients in water would allow for a relatively easier detection of carrion by snakes.
This gained some support from \cite{devault2002scavenging}, who found a preponderance of aquatic snake species in their review of this behaviour.
Smell seems to be the primary means of carcass detection in sharks as well. 
\cite{fallows2013white} found that wind speed determined the number of sharks feeding at whale carcasses, indicating they were dependent on detecting the odours from the decaying whales.

\subsection*{Prey Availability}
The environmental influence on prey availability is an aspect that greatly affects encounter rate but is invisible to the selective forces acting on the scavenger. 
Aspects including, primary productivity, relief, and temperature will all greatly affect scavenging tendency. 
\cite{ruxton2004obligate} suggest an historic ecosystem with a productivity similar to the Serengeti could have supported an \textit{obligate} mammalian or reptilian terrestrial scavenger.
Indeed, in systems that were dominated by large ectothermic or mesothermic herbivore vertebrates, the same primary productivity would have supported a greater biomass, due to the scaling of mass with metabolic rate \citep{mcnab2009resources}.
The upshot of this may have been a higher biomass of herbivores dying and offering scavenging opportunities (although these larger species may have also lived longer).
